---
title: "week_07_lab7: The Bootstrap"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

moths <- read.csv(file = "https://michaelfrancenelson.github.io/eco_602_634_2020/data/moths.csv")

# moths contains standardized abundance estimates of 10 rare moth species across 24 sample plots in the pine barrens of southeast Massachusetts.
```


```{r cars}
hist(moths$anst) # Anisota stigma distribution... not a normal distribution

# what can we do to get a 95% confidence interval?
```


t-distribution: normal distribution when the population variance is unknown and must be estimated from the sample.
t-distributions have heavier tails than normal distributions, so they are a bit more conservative. 
```{r CI interval radius}
alpha = 0.05
anst = moths$anst
n = sum(!is.na(anst))
t_crit = abs(qt(alpha / 2, df = n - 1))

sse = sd(anst) / sqrt(n)

sample_mean = mean(anst)
ci_parametric = sse * t_crit

confidence_intervals = 
  data.frame(
    technique = c("parametric: t-dist"),
    mean = sample_mean,
    ci_radius = sse * t_crit,
    lower = sample_mean - ci_parametric,
    upper = sample_mean + ci_parametric
  )
```
This is the standard way to construct a confidence interval and test the hypothesis that the mean differs from 0. 


We can perform a bootstrap simulation by calculating the mean abundance on many randomly sampled (with replacement) data sets. 
```{r bootstrap}
# 1. create an empty vector to hold the bootstrap sample means
m = 10000
result = numeric(m)
head(result)

# 2. create the resampled data sets and calculate the means

for(i in 1:m)
{
  result[i] = mean(sample(anst, replace=TRUE))
}

# 3. calculate the CI from the quantiles of the resulting bootstrap means

mean(result)
quantile(result,c(0.025,0.975))
```


Bootstrap interval using boot()
```{r something}
install.packages("boot")
library(boot)
# boot(data, statistic, n resamples)

boot_mean = function(x, i) # create a function to calculate mean
{
  return(mean(x[i]))
} 

myboot = 
  boot(
    data = anst,
    statistic = boot_mean,
    R = 10000)
print(myboot)

```


```{r IDK what this means}
mean(anst)
myboot$t0
mean(myboot$t) - myboot$t0
sd(myboot$t)

quantile(
  myboot$t,
  c(0.025, 0.975))

```




```{r something}
moth_dat = moths[,-1]
head(moth_dat)

n = nrow(moth_dat) #number of rows or sample observations
m = 100 #number of bootstrap iterations
moth_result = matrix(
  nrow = m,
  ncol = n)

# The outer loop: runs once for each bootstrap iteration.  index variable is i
for(i in 1:m)
{
  # The inner loop: simulates increasing sampling intensity
  # Sampling intensity ranges from 1 site to the complete count of sites (24)
  # index variable is j
  for(j in 1:n)
  {
    # sample the input data row indices, with replacement
    rows_j = sample(n, size = j, replace=TRUE)
    
    # Creates a new data matrix
    t1 = moth_dat[rows_j, ]
    
    # Calculates the column sums
    t2 = apply(t1, 2, sum)
    
    # Counts the number of columns in which any moths were observed
    moth_result[i, j] = sum(t2 > 0)
  }
}

head(moth_result)
````

```{r first draft}

rarefaction_sampler = function(input_dat, n_iterations)
{
  n = nrow(moth_dat) #number of rows or sample observations
  m = 100 #number of bootstrap iterations
  
  moth_result = matrix(
    nrow = m,
    ncol = n)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:m)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of sites (24)
    # index variable is j
    for(j in 1:n)
    {
      
      # sample the input data row indices, with replacement
      rows_j = sample(n, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = moth_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      moth_result[i, j] = sum(t2 > 0)
    }
  }
  
  return(moth_result)
}

rarefact = rarefaction_sampler(moth_dat, 100)
head(rarefact)
```

```{r second draft}
rarefaction_sampler = function(input_dat, n_iterations)
{
  n_input_rows = nrow(input_dat)
  
  results_out = matrix(
    nrow = n_iterations,
    ncol = n_input_rows)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:n_iterations)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of 
    # sites in the input data (n)
    # index variable is j
    for(j in 1:n)
    {
      # sample the input data row indices, with replacement
      rows_j = sample(n, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = input_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      results_out[i, j] = sum(t2 > 0)
    }
  }
  return(results_out)
}

rarefact = rarefaction_sampler(moth_dat, 100)
head(rarefact)

```

```{r check in a fresh environment}
# This clears the current R session's environment
rm(list = ls())

# Re-read my data:
moths = read.csv(here("data", "moths.csv"))
moth_dat = moths[,-1]

rarefaction_sampler = function(input_dat, n_iterations)
{
  
  n_input_rows = nrow(input_dat)
  
  results_out = matrix(
    nrow = n_iterations,
    ncol = n_input_rows)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:n_iterations)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of 
    # sites in the input data (n)
    for(j in 1:n)
    {
      # sample the input data row indices, with replacement
      rows_j = sample(n, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = input_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      results_out[i, j] = sum(t2 > 0)
    }
  }
  return(results_out)
}

rarefact = rarefaction_sampler(moth_dat, 100)

```

```{r debugging template}
# This clears the current R session's environment
rm(list = ls())

# Re-read my data:
moths = read.csv(here("data", "moths.csv"))

rarefaction_sampler = function(input_dat, n_iterations)
{
  ... paste your corrected code here ...
}

rarefact = rarefaction_sampler(moths[,-1], 100)
head(rarefact)
```


```{r rarefaction curve}
# Re-read my data:
moths = read.csv(file = "https://michaelfrancenelson.github.io/eco_602_634_2020/data/moths.csv")
rarefact = rarefaction_sampler(moths[,-1], 10000)
```



