---
title: "week_07_lab7: The Bootstrap"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

moths <- read.csv(file = "https://michaelfrancenelson.github.io/eco_602_634_2020/data/moths.csv")

# moths contains standardized abundance estimates of 10 rare moth species across 24 sample plots in the pine barrens of southeast Massachusetts.
```


```{r hist}
hist(moths$anst) # Anisota stigma distribution... not a normal distribution

# what can we do to get a 95% confidence interval?
```


t-distribution: normal distribution when the population variance is unknown and must be estimated from the sample.
t-distributions have heavier tails than normal distributions, so they are a bit more conservative. 
```{r CI interval radius using T}
alpha = 0.05
anst = moths$anst # spiny oakworm, Anisota stigma
n = sum(!is.na(anst))
t_crit = abs(qt(alpha / 2, df = n - 1)) # multiply critical t-value by the sample standard error of the mean to get the radius of the CI. 

sse = sd(anst) / sqrt(n)

sample_mean = mean(anst)
ci_parametric = sse * t_crit

confidence_intervals = 
  data.frame(
    technique = c("parametric: t-dist"),
    mean = sample_mean,
    ci_radius = sse * t_crit,
    lower = sample_mean - ci_parametric,
    upper = sample_mean + ci_parametric
  )
confidence_intervals
```
This is the standard way to construct a confidence interval and test the hypothesis that the mean differs from 0. 


We can perform a bootstrap simulation by calculating the mean abundance on many randomly sampled (with replacement) data sets. This is an alternate way to obtain the CI. 
```{r bootstrap}
# 1. create an empty vector to hold the bootstrap sample means
m = 10000
result = numeric(m)
head(result)

# 2. create the resampled data sets and calculate the means

for(i in 1:m)
{
  result[i] = mean(sample(anst, replace=TRUE))
}

# 3. calculate the CI from the quantiles of the resulting bootstrap means

mean(result)
quantile(result,c(0.025,0.975))
```


Bootstrap interval using boot()
```{r something}
install.packages("boot")
library(boot)
# boot(data, statistic, n resamples)

boot_mean = function(x, i) # create a function to calculate mean
{
  return(mean(x[i]))
} 

myboot = 
  boot(
    data = anst,
    statistic = boot_mean,
    R = 10000)
print(myboot)

# original is the original mean of the whole sample (mean(anst))
# bias is the difference between the original mean and the mean of bootstrapped samples
# std.error is the standard deviation of the simulated values
str(myboot)
```


```{r extract bootstrap CI intervals from }
mean(anst)
myboot$t0
mean(myboot$t) - myboot$t0
sd(myboot$t)

quantile(
  myboot$t,
  c(0.025, 0.975)) # this is how to extract the bootstrap confidence intervals

```
So far we have used three different methods to extract CIs: t-distribution sampling, simple boot, and the boot package boot() function. 



```{r setting up the bootstrap}
moth_dat = moths[,-1] # remove siteID column
head(moth_dat)

n = nrow(moth_dat) #number of rows or sample observations
m = 100 # number of bootstrap iterations
moth_result = matrix(
  nrow = m,
  ncol = n) # this is an empty matrix to contain the bootstrap


# The outer loop: runs once for each bootstrap iteration.  index variable is i
for(i in 1:m)
{
  # The inner loop: simulates increasing sampling intensity
  # Sampling intensity ranges from 1 site to the complete count of sites (24)
  # index variable is j
  for(j in 1:n)
  {
    # sample the input data row indices, with replacement
    rows_j = sample(n, size = j, replace=TRUE)
    
    # Creates a new data matrix
    t1 = moth_dat[rows_j, ]
    
    # Calculates the column sums
    t2 = apply(t1, 2, sum)
    
    # Counts the number of columns in which any moths were observed
    moth_result[i, j] = sum(t2 > 0)
  }
}

head(moth_result)
````

```{r first draft of portable rarefaction function}

rarefaction_sampler = function(input_dat, n_iterations)
{
  n = nrow(moth_dat) #number of rows or sample observations
  m = 100 #number of bootstrap iterations
  
  moth_result = matrix(
    nrow = m,
    ncol = n)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:m)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of sites (24)
    # index variable is j
    for(j in 1:n)
    {
      
      # sample the input data row indices, with replacement
      rows_j = sample(n, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = moth_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      moth_result[i, j] = sum(t2 > 0)
    }
  }
  
  return(moth_result)
}

rarefact = rarefaction_sampler(moth_dat, 100)
head(rarefact)
```

```{r second draft portable function}
rarefaction_sampler = function(input_dat, n_iterations)
{
  n_input_rows = nrow(input_dat)
  
  results_out = matrix(
    nrow = n_iterations,
    ncol = n_input_rows)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:n_iterations)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of 
    # sites in the input data (n)
    # index variable is j
    for(j in 1:n_input_rows)
    {
      # sample the input data row indices, with replacement
      rows_j = sample(n_input_rows, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = input_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      results_out[i, j] = sum(t2 > 0)
    }
  }
  return(results_out)
}

rarefact = rarefaction_sampler(moth_dat, 100)
head(rarefact)

```

```{r check in a fresh environment}
# This clears the current R session's environment
rm(list = ls())

# Re-read my data:
moths = read.csv(file = "https://michaelfrancenelson.github.io/eco_602_634_2020/data/moths.csv")
moth_dat = moths[,-1]

rarefaction_sampler = function(input_dat, n_iterations)
{
  
  n_input_rows = nrow(input_dat)
  
  results_out = matrix(
    nrow = n_iterations,
    ncol = n_input_rows)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:n_iterations)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of 
    # sites in the input data (n)
    for(j in 1:n_input_rows)
    {
      # sample the input data row indices, with replacement
      rows_j = sample(n_input_rows, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = input_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      results_out[i, j] = sum(t2 > 0)
    }
  }
  return(results_out)
}

rarefact = rarefaction_sampler(moth_dat, 100)
rarefact
```



```{r debugged rarefaction code: rarefaction_sampler()}
# This clears the current R session's environment
rm(list = ls())

# Re-read my data:
moths = read.csv(file = "https://michaelfrancenelson.github.io/eco_602_634_2020/data/moths.csv")

rarefaction_sampler = function(input_dat, n_iterations)
{
  
  n_input_rows = nrow(input_dat)
  
  results_out = matrix(
    nrow = n_iterations,
    ncol = n_input_rows)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:n_iterations)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of 
    # sites in the input data (n)
    for(j in 1:n_input_rows)
    {
      # sample the input data row indices, with replacement
      rows_j = sample(n_input_rows, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = input_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      results_out[i, j] = sum(t2 > 0)
    }
  }
  return(results_out)
}

rarefact = rarefaction_sampler(moths[,-1], 100)
head(rarefact)
```


```{r rarefaction matrix}
# Re-read my data:
moths = read.csv(file = "https://michaelfrancenelson.github.io/eco_602_634_2020/data/moths.csv")
rarefact = rarefaction_sampler(moths[,-1], 10000)
head(rarefact)
# rarefact will contain 10,000 rows and 24 columns, where each row is a separate bootstrap iteration and each column represents the size of the bootstrap sample (n = 1:24)
```


We can now calculate the mean 2.5 and 97.5% quantiles of the bootstrapped species richness for each sampling intensity using apply()
```{r rarefaction curve}
rare_mean = apply(rarefact, 2, mean)
rare_quant = apply(rarefact, 2, quantile, probs=c(0.025, 0.975))
rare = t(rbind(rare_mean, rare_quant))

matplot(
  rare,
  type='l',
  xlab='Number of sampling plots',
  ylab='Species richness',
  main='Rarefaction Curve')

legend(
  'bottomright',
  legend=c('mean','2.5%','97.5%'),
  lty=c(1,2,3),col=c(1,2,3), inset=c(.1,.1))
```
Five sample plots are not sufficient for the production of confidence intervals. Our CI is tighter at 10 sample plots, and pretty damn tight at 15. 








LAB QUESTIONS
```{r Q1: t-distribution CI}
library(palmerpenguins)
data(package = "palmerpenguins")
penguins <- as.data.frame(penguins)

gentoo <- subset(penguins, penguins$species == "Gentoo")

alpha = 0.05
billlength = gentoo$bill_length_mm[!is.na(gentoo$bill_length_mm)==TRUE]
n = length(billlength)
t_crit = abs(qt(alpha / 2, df = n - 1)) 

sse = sd(billlength) / sqrt(n)

sample_mean = mean(billlength)
ci_parametric = sse * t_crit

confidence_intervals = 
  data.frame(
    technique = c("parametric: t-dist"),
    mean = sample_mean,
    std.dev = sd(billlength, na.rm = TRUE),
    std.err = sse,
    ci_radius = sse * t_crit,
    lower = sample_mean - ci_parametric,
    upper = sample_mean + ci_parametric
  )
print(confidence_intervals)
```
Q1.1 Calculate confidence intervals using student's t-distribution. Paste the above code.
# mean = 47.50488
# sd = 3.081857 
# SE = 0.2778817
# ci_radius = 0.5500946
# lower = 46.95478 
# upper = 48.05497



```{r Q2: boot() CI}
install.packages("boot")
library(boot)
# boot(data, statistic, n resamples)

boot_mean = function(x, i) # create a function to calculate mean
{
  return(mean(x[i]))
} 

myboot = 
  boot(
    data = billlength,
    statistic = boot_mean,
    R = 10000)
print(myboot)


mean(billlength)
myboot$t0
mean(myboot$t) - myboot$t0
sd(myboot$t)

CIbounds <- quantile(
  myboot$t,
  c(0.025, 0.975))
CIbounds
```
Q2.1 Calculate confidence intervals using the boot() function from the package boot(). Paste your code. 
# lower = 46.96992
# upper = 48.05205 


```{r Q3: rarefaction curve}
rm(list = ls())

library(palmerpenguins)
data(package = "palmerpenguins")
penguins <- as.data.frame(penguins)
gentoo <- subset(penguins, penguins$species == "Gentoo")
billlength = as.data.frame(gentoo$bill_length_mm[!is.na(gentoo$bill_length_mm)==TRUE])


rarefaction_sampler = function(input_dat, n_iterations)
{
  n_input_rows = nrow(input_dat)
  
  results_out = matrix(
    nrow = n_iterations,
    ncol = n_input_rows)
  
  # The outer loop: runs once for each bootstrap iteration.  index variable is i
  for(i in 1:n_iterations)
  {
    # The inner loop: simulates increasing sampling intensity
    # Sampling intensity ranges from 1 site to the complete count of 
    # sites in the input data (n)
    for(j in 1:n_input_rows)
    {
      # sample the input data row indices, with replacement
      rows_j = sample(n_input_rows, size = j, replace=TRUE)
      
      # Creates a new data matrix
      t1 = input_dat[rows_j, ]
      
      # Calculates the column sums
      t2 = apply(t1, 2, sum)
      
      # Counts the number of columns in which any moths were observed
      results_out[i, j] = sum(t2 > 0)
    }
  }
  return(results_out)
}

rarefact = rarefaction_sampler(billlength, 1000)
head(rarefact)
```
Q3.1 Paste your completed rarefaction_sampler() function.






















