---
title: "week_06_frequentist_concepts"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R implements probability, cumulative probability, and quantile functions for many parametric distributions. These are represented by a p, d, or q respectively in say, pbinom()

pbinom: cumulative mass function
dbinom: probability mass function
qbinom: quantile function

Examples of their applications: 

Is a value of 1.2 or 2.4 more likely? - The probability density function, e.g. dnorm()

What is the probability that I observe a value between 1.2 and 2.4?: The cumulative density function, e.g. pnorm()

What is the probability that I observe a value of 1.3 or more?: The cumulative density function, e.g. pnorm()

What is the 20th percentile of fish lengths?: The quantile function, e.g. qnorm()

```{r pnorm}
# pnorm() fives the probability of observing a value of x or less
pnorm(7.5, mean = 10, sd = 3) # P 7.5 or less
1-(pnorm(7.5, mean = 10, sd = 3)) # P 7.5 or more
```


```{r Q1}
dbinom(3, 4, 0.75)
```
Q1.1 What is the probability of observing a count of exactly 3 successes in a binomial distribution with parameters n = 4 and p = 0.75?
P = 0.421875

Q1.2 Paste the R code that you used to find your answer.
dbinom(3, 4, 0.75)


```{r Q2}
pbinom(3, 4, 0.75)
```
Q2.1 What is the probability of observing a count of 3 successes or fewer in a binomial distribution with parameters n = 4 and p = 0.75?
P = 0.6835937

Q2.1 Paste the R code that you used to find your answer.
pbinom(3, 4, 0.75)


```{r Q3}
1-pbinom(3, 5, 0.75)
sum(dbinom(4:5, 5, 0.75))
```
Q3.1 What is the probability of observing more than 3 successes in a binomial distribution with parameters n = 5 and p = 0.75?
P = 0.6328125

Q3.2 Paste the R code that you used to find your answer.
1-pbinom(3, 5, 0.75)


```{r Q4}
pnorm(1.2, mean = 2, sd = 2) 
```
Q4.1 What is the probability of observing a value of less than 1.2 from a normally-distributed population with mean = 2 and standard deviation = 2?
P = 0.3445783

Q4.2 Paste the R code that you used to find your answer.
pnorm(1.2, mean = 2, sd = 2) 


```{r Q5}
1 - pnorm(1.2, mean = 2, sd = 2)
```

Q5.1 What is the probability of observing a value of greater than 1.2 from a normally-distributed population with mean = 2 and standard deviation = 2?
P = 0.6554217

Q5.2 Paste the R code that you used to find your answer.
1 - pnorm(1.2, mean = 2, sd = 2) 


```{r Q6}
pnorm(3.2, mean = 2, sd = 2) - pnorm(1.2, mean = 2, sd = 2)
```
Q6.1 What is the probability of observing a value between 1.2 and 3.2 from a normally-distributed population with mean = 2 and standard deviation = 2?
P = 0.3811686

Q6.2 Paste the R code that you used to find your answer.
pnorm(3.2, mean = 2, sd = 2) - pnorm(1.2, mean = 2, sd = 2)



Q7 Describe how the shape of the histogram changes as you continue to press the sample button.
The shape of the histogram approaches that of the original population distribution as more samples of n=1 are drawn. 


Q8 Describe how the shape of the histogram changes as you continue to press the sample button.
The shape of the histogram appears to be somewhere in between the shape of the original distribution and the shape of a normal distribution as more samples of n=2 are drawn. There are fewer means with values close to the tail end of the distribution, unlike when the same activity was performed with samples of n=1. 


Q9 Describe how the shape of the histogram changes as you continue to press the sample button.
The shape of the histogram continues to approach a normal distribution, with a much narrower distribution than that of the distribution derived from the means of samples of n=2. This mirrors what we learned about confidence intervals. The larger the sample size, the smaller the standard deviation about the mean. 

