---
title: "Week 3 Lab: Data Exploration & Deterministic Functions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("psych")
library(psych)
pairs.panels(iris)
write.csv(read.csv("https://michaelfrancenelson.github.io/eco_602_634_2020/data/bird.sta.csv"),"./data//bird.sta.csv", row.names = FALSE) 
dat_bird <- read.csv("./data//bird.sta.csv", sep = ",")
dat_habitat <- read.csv("./data//hab.sta.csv", sep = ",")

```

```{r merge}
dat_bird[colnames(dat_bird)==colnames(dat_habitat)]
colnames(dat_bird)
# shared colnames = basin, sub, sta
dat_all <- merge(dat_habitat, dat_bird)
plot(ba.tot ~ elev, data = dat_all) # test scatterplot
```

```{r binary data}
# it may be useful to convert data with lots of 0s to presence/absence data (Boolean vectors, in essence)
cewa_present_absent <- as.numeric(dat_all$CEWA!=0)
# convert numeric data to boolean vector (TRUE > 0, FALSE = 0), then convert back to numeric data - thus producing binary values of 0s and 1s. 
plot(x = dat_all$elev, y = cewa_present_absent)

```

Visually estimating the parameters of a logistic curve...
  - You can slide the curve to the left or right (via midpoint)
  - You can make the curve steeper or shallower near its midpoint (i.e. slope)

Here's a set of functions for visually parameterizing a logistic function.
```{r logistic parameterization}
# Function to calculate the logistic parameter a given the slope and midpoint
get_logistic_param_a = function(slope, midpoint)
{
  b = slope / 4
  return (-midpoint * (slope / 4))
}

# Function to calculate the logistic parameter b given the slope
get_logistic_param_b = function(slope)
{
  return (slope / 4)
}


# Calculate the value of the logistic function at x, given the parameters a and b.
logistic = function(x, a, b)
{
  val = exp(a + b * x)
  return(val / (1 + val))
}

# Calculate the value of the logistic function at x, given a slopoe and midpoint.
logistic_midpoint_slope = function(x, midpoint, slope)
{
  b = get_logistic_param_b(slope)
  a = get_logistic_param_a(slope, midpoint)
  return(logistic(x, a, b))
}
```


```{r logistic tests}
plot(x = dat_all$elev, y = cewa_present_absent)
curve(logistic_midpoint_slope(x, midpoint = 400, slope = 0.1), add = TRUE)

plot(x = dat_all$elev, y = cewa_present_absent)
curve(logistic_midpoint_slope(x, midpoint = 400, slope = -0.1), add = TRUE)

plot(x = dat_all$elev, y = cewa_present_absent)
curve(logistic_midpoint_slope(x, midpoint = 400, slope = -0.05), add = TRUE)
```

Inserting code chunk for custom pairplot functions
```{r Kristen's pairplot functions}
panel.hist <- function(x, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(usr[1:2], 0, 1.5) )
  h <- hist(x, plot = FALSE)
  breaks <- h$breaks; nB <- length(breaks)
  y <- h$counts; y <- y/max(y)
  rect(breaks[-nB], 0, breaks[-1], y, col="cyan", ...)
}

panel.cor <- function(x, y, digits=2, cex.cor)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits=digits)[1]
  test <- cor.test(x,y)
  Signif <- ifelse(round(test$p.value,3)<0.001,"p<0.001",paste("p=",round(test$p.value,3)))  
  text(0.5, 0.25, paste("r=",txt), cex=1.3)
  text(.5, .75, Signif, cex=1.2)
}
```

```{r terrain pairs}
terrain <- dat_all[, c("elev", "slope", "aspect", "ba.tot")]

pairs(terrain, lower.panel = panel.smooth, upper.panel = panel.cor, diag.panel = panel.hist)
```
There are no substantial relationships between terrain variables and total basal area (despite a small yet significant positive relationship between aspect and ba.tot)


```{r logistic model 1 - CEWA}
# basal area on the x axis, bird presence/absence eon the y-axis (two species)

cewa_present_absent <- as.numeric(dat_all$CEWA!=0)
# convert numeric data to boolean vector (TRUE > 0, FALSE = 0), then convert back to numeric data - thus producing binary values of 0s and 1s. 
plot(x = dat_all$ba.tot, y = cewa_present_absent, main = "Cedar Waxwing Presence vs. Total Basal Area", xlab = "Total basal area (sq.m/ha)", ylab = "Cedar waxwing presence")
curve(logistic_midpoint_slope(x, midpoint = 50, slope = -0.6), add = TRUE)
```
Question 1: Logistic model 1
Consider the first bird species you chose to examine.
Q1.1 (1 pt). Upload an image file of the logistic function plot.
Q1.2 (3 pts). What was the bird species? Describe its presence/absence patterns in terms of basal area. Did it seem to prefer areas with high or low tree cover?
The bird species considered in this plot was the Cedar Waxwing. Cedar Waxwings appear to be present only in areas with low total basal area, below values of about 50 square meters per hectare. Based on its abundance only in study sites with low total basal area, it appears that this bird species prefers areas with low tree cover. 


```{r logistic model 2 - WREN}
wren_present_absent <- as.numeric(dat_all$WREN!=0)
# convert numeric data to boolean vector (TRUE > 0, FALSE = 0), then convert back to numeric data - thus producing binary values of 0s and 1s. 
plot(x = dat_all$ba.tot, y = wren_present_absent, main = "Wrentit Presence vs. Total Basal Area", xlab = "Total basal area (sq.m/ha)", ylab = "Wrentit presence")
curve(logistic_midpoint_slope(x, midpoint = 60, slope = -0.6), add = TRUE)
```
Question 2: Logistic model 2
Consider the second bird species you chose to examine.
Q2.1 (1 pt). Upload an image file of the logistic function plot.
Q2.2 (3 pts). What was the bird species? Describe its presence/absence patterns in terms of basal area. Did it seem to prefer areas with high or low tree cover?
The bird species considered in this plot was the Wrentit. Wrentits appear to be present only in areas with low total basal area, below values of about 60 square meters per hectare. Based on its abundance only in study sites with low total basal area, it appears that this bird species prefers areas with low tree cover. 


```{r jays count}
sum(dat_all$GRJA, na.rm = TRUE)
```
Question 3: Counting total jays
Using the dat_all data frame, calculate the total number of Gray Jays observed in all of the sampling sites.
Q3.1 (1pt): Past the R code you used for the calculation into the text entry window.

sum(dat_all$GRJA, na.rm = TRUE)
There were 181 observations of Gray Jays.


```{r jays presence count}
sum(as.numeric(dat_all$GRJA!=0))
```

Question 4: Counting total sites with jays
Using the dat_all data frame, calculate the total number of sampling sites in which Gray Jays were observed.
Hint: What happens when you use the sum() function on a vector of Boolean values?
Q4.1 (1pt): Past the R code you used for the calculation into the text entry window.

sum(as.numeric(dat_all$GRJA!=0))
There were 110 sampling sites in which Gray Jays were observed.