---
title: "week_06_lab6: Introduction to Inference"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)

sse_mean <- function(x){
  sd(x, na.rm = TRUE)/(sqrt(length(!is.na(x))))
}

sse_mean(penguins$bill_depth_mm)
```


p-value interpretation: "how often would I expect to see a result as extreme, or more extreme, if the null hypothesis were true?"

We can use ANOVA (analysis of variance) to perform a parametric, frequentist analysis of the difference between two groups of values. 
```{r penguins boxplots 1}
boxplot(flipper_length_mm ~ species, data = penguins)

# droplevels() can be used to remove unused factors from a data.frame.
dat_pen = droplevels(subset(penguins, species != "Gentoo"))
{
  par(mfrow = c(1, 2))
  boxplot(flipper_length_mm ~ species, data = penguins)
  boxplot(flipper_length_mm ~ species, data = dat_pen)
}
```


Alternatively, we can do a resampling simulation to estimate how likely the observed differences in flipper length would be if there truly was no difference. Frequentists love it. 
```{r repeated resampling}
# for reproduceability
set.seed(123)

flipper_shuffled = sample(penguins$flipper_length_mm, replace = TRUE)
par(mfrow = c(1, 2))
boxplot(flipper_length_mm ~ species, data = penguins)
boxplot(flipper_shuffled ~ penguins$species, xlab = "species")
# simulated data has much lower variability than the original data grouped by species. 
```


The two-sample t-test is a good way to compare the mean values of two groups.
Two-sample resampling and t-test on simulated data.
```{r two-sample resampling}
set.seed(1)
flipper_shuffled = sample(dat_pen$flipper_length_mm, replace = TRUE)
boxplot(flipper_shuffled ~ dat_pen$species)

t_test_1 = t.test(flipper_shuffled ~ dat_pen$species)
t_test_1
# non-significant t-test: resampled data do not differ significantly, unlike observed data. 
```


Difference of means: 
```{r mean comparison}
t_test = t.test(dat_pen$flipper_length_mm ~ dat_pen$species)
t_test
t_test$estimate
# tiny p value = groups are significantly different in observed data

# difference in means found below
diff_observed = round(diff(t_test$estimate), digits = 3)
print(diff_observed, digits = 3)
# = 5.87

```

The difference in means can also be found by using the aggregate() function.
```{r aggregate means}
# aggregate() splits the data into subsets, computes summary statistics for each, and returns the result in a convenient form.
agg_means = aggregate(
  flipper_length_mm ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)
diff_observed = diff(agg_means[, 2])

agg_means
diff_observed
```


Sample sizes
```{r sample sizes}
table(dat_pen$species) # provides number of values in each category
# adelie = 152; chinstrap = 68
```

Resampling with replacement is the same thing as randomly sampling 68 flipper lengths in one group and 152 in another
```{r resampling with replacement}
n_1 = 68
n_2 = 152

dat_1 = sample(dat_pen$flipper_length_mm, n_1, replace = TRUE)
dat_2 = sample(dat_pen$flipper_length_mm, n_2, replace = TRUE)

diff_simulated = 
  mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)

print(c(observed = diff_observed, simulated = diff_simulated))
# the difference between groups is very different between observed and simulated data. observed diff = 5.87, simulated = 0.69.
```

This simulation can be consolidated into a single function.
```{r two-group resample function}
two_group_resample = function(x, n_1, n_2){
  dat_1 = sample(x, n_1, replace = TRUE)
  dat_2 = sample(x, n_2, replace = TRUE)
  diff_simulated = mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)
  print(diff_simulated)
}
```

```{r testing custom resample function}
set.seed(54321)
two_group_resample(dat_pen$flipper_length_mm, 68, 152)
```


If I ran this resampling function a bunch of times, how often would I see a mean difference greater than diff_observed?
```{r repeated resampling}
n = 2000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$flipper_length_mm, 68, 152)
  )
}
hist(mean_differences)

# no simulation differences were found to be greater than the observed difference in means. This is consistent with the t-test.
```


Retrieving named elements
```{r retrieving named elements}
t_test = t.test(flipper_shuffled ~ dat_pen$species)
str(t_test)
t_test$estimate

# this way, we can retrieve the desired elements from a particular statistical test. in this case, we isolated the mean values of Adelie and Chinstrap penguins. 
```




LAB QUESTIONS


Question 01: Standard error of the mean function
Q1.1 Paste the code you used to define your sse_mean() function into the text input.

sse_mean <- function(x){
  sd(x, na.rm = TRUE)/(sqrt(length(!is.na(x))))
}



Question 02: Resampling
Q2.1 Paste the code you used to define your two_group_resample() function into the text input.

two_group_resample = function(x, n_1, n_2){
  dat_1 = sample(x, n_1, replace = TRUE)
  dat_2 = sample(x, n_2, replace = TRUE)
  diff_simulated = mean(dat_1, na.rm = TRUE) - mean(dat_2, na.rm = TRUE)
  print(diff_simulated)
}



Question 03: Resampling
```{r Q4}
n = 2000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$flipper_length_mm, 68, 152)
  )
}
hist(mean_differences, main = "Histogram of Differences in Repeatedly\n Resampled Flipper Length Means", xlab = "Difference in mean (mm)")
```
Q3.1 Upload a histogram of the resampled differences of means.

lab_06_mean_difs_hist FIGURE



Question 04: Resampling
```{r Q4}
n = 2000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$flipper_length_mm, 68, 152)
  )
}

sum(abs(mean_differences)>5.8)
```
Q4.1 How many of your resampled differences of means were greater than 5.8? Assume we are doing a 2-sided test (we don't care which species has longer flippers).

0 resampled differences of means were greater than 5.8. 

Q5.2 Paste the R code you used to check into the text entry box.

n = 2000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$flipper_length_mm, 68, 152)
  )
}
sum(abs(mean_differences)>5.8)



Question 05: Resampling
Q5.1 Given a p value of less than 1 per 10 million, how many simulations do you think you would have to do to see a difference in mean flipper length equal to or greater than 5.8 mm?

It would take an average of 16,531,658 simulations (1 divided by the p-value) to see a difference in mean flipper length equal to or greater than 5.8 mm.



Question 06: Resampling
```{r Q6}
# "bill_length_mm"

# boxplot of measurements, grouped by species
dat_pen = droplevels(subset(penguins, species != "Gentoo"))
boxplot(bill_length_mm ~ species, data = dat_pen, main = "Bill Length by Penguin Species", xlab = "Species", ylab = "Bill length (mm)")

# use aggregate() to calculate the group means and differences in means, and write the difference in means into a variable called diff_crit
agg_means = aggregate(
  bill_length_mm ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)
diff_crit = diff(agg_means[, 2])

# conduct a t-test and observe p-value
t_test = t.test(dat_pen$bill_length_mm ~ dat_pen$species)
t_test
# p-value < 2.2e-16

# conduct a resampling test with 1000 repetitions using two_group_resample()
tinyguin <- subset(dat_pen, !is.na(dat_pen$bill_length_mm))
table(tinyguin$species) # 151 Adelie, 68 Chinstrap

n = 1000
mean_differences = c()
for (i in 1:n)
{
  mean_differences = c(
    mean_differences,
    two_group_resample(dat_pen$bill_length_mm, 68, 151)
  )
}



```
Q6.1 Upload an image file of the boxplot of your chosen variable.

lab_06_bill_byspecies_box FIGURE 

Q6.2 What were the group means and difference between the means? Paste your answer and the R code you used into the text entry box.

The mean bill length (mm) is 38.791 for Adelie and 48.834 for Chinstrap penguins. The difference in these means is 10.04243.

agg_means = aggregate(
  bill_length_mm ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)
diff_crit = diff(agg_means[, 2])





Question 07: Resampling
Q7.1 Interpret the p-value from the t-test output in plain, non-technical english that a non-statistician would understand.

The p-value determined by a t-test comparing the mean bill length of Adelie to Chinstrap penguins is < 2.2e-16. This means that there is an incredibly small probability that the bill lengths of Adelie and Chinstrap penguins differ solely by chance. In other words, there is a significance difference between the bill lengths of these two species. 




Question 08: Resampling
```{r Q8}
agg_means = aggregate(
  bill_length_mm ~ species, 
  data = dat_pen, 
  FUN = mean, 
  na.rm = TRUE)
diff_crit = diff(agg_means[, 2])

sum(abs(mean_differences)>abs(diff_crit))

hist(mean_differences, main = "Histogram of Differences in Repeatedly\n Resampled Bill Length Means", xlab = "Difference in mean (mm)")

```
Q8.1 How many differences in means were greater than diff_crit. Remember to treat this as a 2-tailed test.

sum(abs(mean_differences)>abs(diff_crit)) = 0.
Zero differences in means were greater than diff_crit. 

Q8.2 Upload a histogram of your simulation results.

lab_06_mean_diffs_hist2 FIGURE




