---
title: 'Lab 5: Modeling Error'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r ricker_fun and sample plot}
ricker_fun = function(x, a, b) 
{
  return(a * x * exp(-b * x))
}


curve(
  ricker_fun(x, 1, 1), 
  from = 0, to = 5, add = FALSE, # from and to indicate x values
  main = "Ricker function: a = 1, b = 1",
  ylab = "f(x)", xlab = "x")
```

```{r exp_fun and sample plot}
exp_fun <- function(x, a, b){
  return(a * exp(-b * x))
}

curve(
  exp_fun(x, 0.1, 0.5), 
  from = 0, to = 10, add = FALSE, 
  main = "Exponential function",
  ylab = "f(x)", xlab = "x")
```

We will use curve() to plot deterministic models using the different functions we defined above.

We can also create simulated data that uses these functions as a deterministic model and creates random deviates to generate noise. 
```{r simulated data on a line}
# Seed the RNG so we can reproduce our results
set.seed(1234567)

# Specify the x-range and number of points:
n_pts = 50
x_min = 2
x_max = 10

# Generate the x-values
x_sim = runif(n_pts, min = x_min, max = x_max)

# runif() generates random deviates on a uniform distribution from min to max. 

# choose an intercept and slope for the deterministic model and generate the predicted y values
param_intercept = 2.3
param_slope = 0.67
y_pred = param_intercept + x_sim * param_slope
plot(x_sim, y_pred)

```

Create observed y values by adding normally distributed noise to predicted values. 
```{r normal errors 1}
error_mean = 0
error_sd = 0.25

y_observed = 
  y_pred + 
  rnorm(
    n = n_pts, 
    mean = error_mean, 
    sd = error_sd)

plot(x_sim, y_observed)
```

Create observed y variables with a greater error about the model. 
```{r normal errors 2}
error_mean = 0
error_sd = 0.1

y_observed_2 = 
  y_pred + 
  rnorm(
    n = n_pts, 
    mean = error_mean, 
    sd = error_sd * x_sim)

plot(x_sim, y_observed_2)
```

Create a model with exponentially-distributed errors using rexp()
```{r exponentially-distributed errors}
rexp(50, rate = 1.2)

y_observed_3 = y_pred + rexp(n = n_pts, rate = 1.2)

plot(x_sim, y_observed_3)
```

```{r choosing a plot}
# View models side by side
par(mfrow = c(3, 1))
plot(x_sim, y_observed)
plot(x_sim, y_observed_2)
plot(x_sim, y_observed_3)


# Examine a histogram of the residuals
par(mfrow = c(3, 1))
hist(y_observed - y_pred, main = "sim data 1", xlab = "observed y=values")
hist(y_observed_2 - y_pred, main = "sim data 2", xlab = "observed y=values")
hist(y_observed_3 - y_pred, main = "sim data 3", xlab = "observed y=values")
```


MARBLED SALAMANDER DISPERSAL DATA
```{r data import}
dat_dispersal <- read.csv("https://michaelfrancenelson.github.io/eco_602_634_2020/data/salamander_dispersal.csv", header = TRUE, sep = ",")
dev.off()
```

```{r examine data}
plot(dispersal$dist.class, dispersal$disp.rate.ftb, main = "Dispersal rates of first time \nbreeder salamanders by distance", xlab = "Distance class (m)", ylab = "Dispersal rate")
# function(x, a, b)

curve(
  ricker_fun(x, .02, .008), # a = slope of initial curve, 1/b = height
  from = 0, to = 1500, add = TRUE,
  main = "Ricker function",
  ylab = "f(x)", xlab = "x", col = "green")

curve(
  exp_fun(x, 1, .003), 
  from = 0, to = 1500, add = TRUE, 
  main = "Exponential function",
  ylab = "f(x)", xlab = "x", col = "blue")

lm(dispersal$disp.rate.ftb ~ dispersal$dist.class)

abline(0.4482125, -0.0003128, col = "red")

```

```{r exponential comparison}
curve(
  exp_fun(x, 1.9, 0.1), 
  from = 0, to = 40, add = FALSE, 
  main = "Exponential function variants",
  ylab = "f(x)", xlab = "x", col = "black", lty = 1)

curve(
  exp_fun(x, 1.9, 0.3), add = TRUE, col = "black", lty = 3)

curve(
  exp_fun(x, 1.2, 0.2), add = TRUE, col = "red", lty = 1)

curve(
  exp_fun(x, 1.2, 0.4), add = TRUE, col = "red", lty = 3)
```

```{r ricker comparison}
curve(
  ricker_fun(x, 25, 0.1), 
  from = 0, to = 40, add = FALSE,
  main = "Ricker function variants",
  ylab = "f(x)", xlab = "x", lty = "solid")
curve(ricker_fun(x, 20, 0.2), add = TRUE, lty = "dotted")
curve(ricker_fun(x, 10, 0.2), add = TRUE, lty = "dotted")
curve(ricker_fun(x, 75, 0.3), add = TRUE, lty = "solid")
curve(ricker_fun(x, 50, 0.3), add = TRUE, lty = "dotted")
curve(ricker_fun(x, 40, 0.3), add = TRUE, lty = "dotted")
```



Question 9: using the parameters chosen for the linear fit, calculate the residuals and store them in a vector "resids_linear"
```{r linear residuals}
lin_pred = 0.4482125 + dispersal$dist.class * -0.0003128
plot(dispersal$dist.class, lin_pred)

resids_linear <- dispersal$disp.rate.ftb - lin_pred
 
hist(resids_linear, main = "Residuals of a linear model of \ndistance vs. salamander dispersal rates", xlab = "Observed y values") 
```

```{r exponential residuals}
exp_pred = exp_fun(dispersal$dist.class, 1, .003)
plot(dispersal$dist.class, exp_pred)
resids_exp <- dispersal$disp.rate.ftb - exp_pred
 
hist(resids_exp, main = "Residuals of an exponential model of \ndistance vs. salamander dispersal rates", xlab = "Observed y values") 
```

```{r ricker residuals}
rick_pred = ricker_fun(dispersal$dist.class, .02, .008)
plot(dispersal$dist.class, rick_pred)
resids_ricker <- dispersal$disp.rate.ftb - rick_pred

hist(resids_ricker, main = "Residuals of a ricker model of \ndistance vs. salamander dispersal rates", xlab = "Observed y values") 
```

